---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# xmlwriter

<!-- badges: start -->
[![CRAN status](https://www.r-pkg.org/badges/version/xmlwriter)](https://CRAN.R-project.org/package=xmlwriter)
[![R-CMD-check](https://github.com/edwindj/xmlwriter/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/edwindj/xmlwriter/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

`xmlwriter` is an R package that provides a simple interface for creating XML documents 
and fragments from R. It has a feed-forward API that allows you to write xml 
in the same order as it appears in an xml document or fragment.
Its main goal is to provide a simple and fast way to create xml documents and fragments.

`xmlwriter` can be used as a companion to R package [`xml2`](https://cran.r-project.org/package=xml2), 
which is a wonderful package optimized for parsing,
querying and manipulating XML documents. `xml2` provides several ways for creating
xml documents, but it is not optimized for generating and writing xml.

Creating xml documents with `xml2` can be a bit cumbersome, because it mostly 
forces the author to manipulate the xml document tree adding nodes and attributes.
It _does_ provide a way to create xml documents from R data structures using
nested lists. This is a very powerful feature, but not optimized for speed
and readability.

`xmlwriter` provides a more natural interface for creating xml documents, 
that mimicks the way xml is written in a text editor.
It has two different ways to create xml documents:

- a light weight R syntax using `xml_fragment` and `.elem` functions, creating 
an `xml_fragment`, that can be easily translated into a xml string or 
`xml2::xml_document` object, or be used as a flexible building block for generating
a larger xml document.
- an `xmlbuilder` object that allows you to create xml documents in a 
feed-forward manner, with `start` and `end` methods, giving you more control on the xml document structure, including xml comment, prolog etc.

## Installation

You can install the development version of xmlwriter from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("edwindj/xmlwriter")
```

## Example

### Using `xml_fragment`, `.elem` functions:

```{r fragment}
library(xmlwriter)

fragment <- xml_fragment(
  persoon = .elem(
    .attr = c(id = "1"),
    name = "John Doe",
    age = 30,
    address = .elem(
       street = "123 Main St",
       city = "Anytown",
       state = "CA",
       zip = 12345
    )
  )
)

print(fragment)
fragment |> xml2::as_xml_document()
fragment |> as_xml_nodeset()
```


### Using `xmlbuilder` object

```{r xmlbuilder}
library(xmlwriter)

b <- xmlbuilder(allow_fragments = FALSE)
b$comment("This is an xml comment")
b$start("person", id = "1")
  b$element("name", "John Doe")
  b$element("age", 30)
  b$start("address")
    b$element("street", "123 Main St")
    b$element("city", "Anytown")
    b$element("state", "CA")
    b$element("zip", 12345)
  b$end()
b$start("person", id = "1")
  b$element("name", "John Doe")
  b$element("age", 30)
  b$start("address")
    b$element("street", "123 Main St")
    b$element("city", "Anytown")
    b$element("state", "CA")
    b$element("zip", 12345)
  b$end()
b$end()

# includes a xml prolog and comment
b$to_xml_string()

# only contains the actual nodes
xml2::as_xml_document(b)
```


# Performance

`xmlwriter` is optimized for generating xml documents and fragments from R data structures.

```{r performance, eval=TRUE}
library(microbenchmark)

library(xml2)
library(xmlwriter)

# read in a sample 600k xml file as a R list str
doc <- xml2::read_xml("./example/DataGeneric.xml")
doc_list <- xml2::as_list(doc)

# copy of the list with an extra attribute class="xml_fragment" (xmlwriter specific)
doc_list2 <- structure(doc_list, class = "xml_fragment")

# see how long it takes to create an xml document with xml2 and xmlwriter

microbenchmark(
  xml2 = xml2::as_xml_document(doc_list),
  xmlwriter = xml2::as_xml_document(doc_list2),
  times = 10
)

```
