---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# xmlwriter

<!-- badges: start -->
[![CRAN status](https://www.r-pkg.org/badges/version/xmlwriter)](https://CRAN.R-project.org/package=xmlwriter)
[![R-CMD-check](https://github.com/edwindj/xmlwriter/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/edwindj/xmlwriter/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

xmlwriter is a package that provides a simple interface for creating XML documents 
and fragments from R. It has a feed-forward API that allows you to write xml 
in the same order as it appears in the document or fragment.

It can be used as a companion to `xml2`, which is a wonderful package optimized for parsing,
querying and manipulating XML documents. `xml2` provides several ways in creating
xml documents, but it is not optimized for writing and creating xml.

Creating xml documents with `xml2` can be a bit cumbersome, because it forces
the author to manipulate the xml document tree adding nodes and attributes in 
a recursive manner.
`xmlwriter` provides a more natural interface for creating xml documents, that
mimicks the way xml is written in a text editor. 
`xmlwriter` is built on top of `Rcpp` and is much faster than `xml2` when
creating xml.


## Installation

You can install the development version of xmlwriter from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("edwindj/xmlwriter")
```

## Example

This is a basic example which shows you how to solve a common problem:

```{r example}
library(xmlwriter)

b <- xmlbuilder()

b$start("person", id = "1")
  b$element("name", "John Doe")
  b$element("age", 30)
  b$start("address")
    b$element("street", "123 Main St")
    b$element("city", "Anytown")
    b$element("state", "CA")
    b$element("zip", 12345)
  b$end()
b$end()

b$to_xml_string()
xml2::as_xml_document(b)
```


# Performance

The package is optimized for generating xml.

```{r performance, eval=TRUE}
library(microbenchmark)
library(xml2)
library(xmlwriter)

# read in a sample 600k xml file as a R list str
doc <- xml2::read_xml("./example/DataGeneric.xml")
doc_list <- xml2::as_list(doc)

# see how long it takes to create an xml document with xml2 and xmlwriter

microbenchmark(
  xml2 = xml2::as_xml_document(doc_list),
  xmlwriter = xmlwriter::list_to_xml_node(doc_list),
  times = 10
)

```
