---
output: github_document
editor_options: 
  markdown: 
    wrap: 72
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# xmlwriter

<!-- badges: start -->

[![CRAN
status](https://www.r-pkg.org/badges/version/xmlwriter)](https://CRAN.R-project.org/package=xmlwriter)
[![R-CMD-check](https://github.com/edwindj/xmlwriter/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/edwindj/xmlwriter/actions/workflows/R-CMD-check.yaml)

<!-- badges: end -->

`xmlwriter` is an R package that provides a simple interface for
creating XML documents and fragments from R. It provides a simple
elegant syntax for creating `xml_fragment`s. Furthermore it contains a
feed-forward API that allows you to write xml in the same order as the
xml elements appear in an xml document or fragment.

`xmlwriter`'s xml generation from R lists is fast, implemented in C++
using [`Rcpp`](https://cran.r-project.org/package=Rcpp). Curious for the
benchmarks? Check the [performance section](#performance).

`xmlwriter` can be used as a companion to R packages
[`XML`](https://cran.r-project.org/package=XML) or
[`xml2`](https://cran.r-project.org/package=xml2) which are both
wonderful packages optimized for parsing, querying and manipulating XML
documents. Both `XML` and `xml2` provide several ways for creating xml
documents, but they are not optimized for generating and writing xml.

Creating xml documents with `XML` and `xml2` can be a bit cumbersome,
mostly because it\
forces the author to manipulate the xml document tree, obscuring the xml
structure of the document, and making it hard to read the code. `xml2`
*does* provide a way to create xml documents from R data structures
using nested lists which is a powerful feature, but it is not optimized
for speed or readability.

`xmlwriter` provides an intuitive interface for creating xml documents,
that mimicks how xml is written in a text editor.

It has two different ways to create xml documents:

**NOTE the api is still in flux, and might change before `xmlwriter` is
put on CRAN**

-   a light weight R syntax using `xml_fragment`, `tag`, `frag` and
    `data_frag`, creating an `xml_fragment`, that can be easily
    translated into a `character` or `xml2::xml_document` object, or be
    used as a flexible building block for generating a larger xml
    document.
-   an `xmlbuilder` object that allows you to create xml documents in a
    feed-forward manner, with `start` and `end` methods, giving you more
    control on the xml document structure, including xml comment, prolog
    etc.

## Installation

`xmlwriter` is not yet available on CRAN, feedback is welcome before
submitting to CRAN.

You can install the development version of `xmlwriter` from
[GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("edwindj/xmlwriter")
```

## Example

### Using `xml_fragment`, `tag`, `frag` and `.attr`:

`xml_fragment`s allow to write XML using readable R syntax.

They can be used to create an `character` with valid xml, a
`xml2::xml_document` or as a building block for more complex XML
documents. An `xml_fragment` is a list object that is identical to the
output of `xml2::as_list`, and can be converted to a `character` or an
`xml2::xml_document` object but is much faster (see performance).

`tag` is a function that creates a simple `xml_fragment` element with a
given tag name, text, and attributes.

```{r}
library(xmlwriter)

# a tag is a simple xml_fragment
tag("person", "John Doe", id = 1)
```

`frag` allows for specifying nested tags, and attributes, creating more
complex xml fragments.

Each argument to `frag` is either:

-   a named `frag` element in which case the name is used as the tag
    name.
-   an unnamed element in which case the element is added as a text
    node.
-   a `.attr` argument that is used to add attributes to the parent
    element.

```{r}
# a frag can contain multiple tags
frag(
  name = "John Doe",
  age = 30
)

# or can nest tags
frag(
  person = frag(
    # attributes are specified using .attr
    .attr = c(id = 1),
    name = "John Doe",
    age = 30
  )
)
```

The `xml_fragment` function is a restricted version of `frag` that only
allows named elements.

The output of `xml_fragment` is a

```{r}
# xml_fragment is more strict version of a frag
fragment <- xml_fragment(
  person = frag(
    .attr = c(id = 1),
    name = "John Doe",
    age = 30,
    address = frag(
      street = "123 Main St",
      city = "Anytown",
      state = "CA",
      zip = 12345
    )
  )
)

```

```{r, eval=FALSE}
cat(as.character(fragment))
```

```{r, results='asis', echo=FALSE}
cat("```XML\n")
cat(as.character(fragment))
cat("\n```\n")
```

`data_frag` is function that can be used to convert a data.frame to an
`xml_fragment`:

```{r data}
data <- data.frame(
  name = c("John Doe", "Jane Doe"),
  age = c(30, 25),
  stringsAsFactors = FALSE
)

# create an xml_fragment from a data.frame
data_frag(data, row_tag = "person")

# but you can also use it within an xml_fragment

# xml_doc is a xml_fragment that contains a single root element
doc <- xml_doc(
  homeless = frag(
    .attr = c(year = "1900"),
    data = data_frag(data, row_tag = "person")
  )
)

doc
```

#### Combine fragments with `+`, `append` or `c()`)

`xml_fragment`s such as `tag`, `frag` and `data_frag` can be combined
with the `+` operator, which is equivalent to the `append()` and `c()`
function: it creates sibling xmlnodes.

```{r}
library(xmlwriter)
john <- tag("person", "John", id = 1)
jane <- tag("person", "Jane", id = 2)

john + jane

john + tag("person", "Jane", id = 2)

john + xml_fragment(
  person = frag(
    .attr = c(id = 2),
    "Jane"
  )
)
```

#### Add child fragments with '/' or `add_child_fragment()`

-   the `/` operator, which is equivalent to the `add_child_fragment`
    function which creates a child xmlnodes of the last xmlnode in an
    `xml_fragment`.

```{r}
tag("person", id = 1) / (
  tag("name", "John Doe") +
  tag("age", 30)
)

tag("person", id = 1) / frag(
  name = "John Doe",
  age = 30
)

tag("person", id = 1) |> 
  add_child_fragment(
    name = "John Doe",
    age = 30
  )
```

#### flexible xml creation...

Using the `+` and `/` operators, one can create complex xml documents in
a very flexible manner.

```{r}
fragment <- 
  tag("person", id = "1") /
    ( tag("name", "John Doe") +
      tag("age", 30) +
      tag("address") /
        frag(
          street = "123 Main St",
          city = "Anytown",
          state = "CA",
          zip = 12345
     )
  )

fragment
```

#### `xml2` compatibility

`xml_writer` does not have a hard dependency on `xml2`, but the output
of xml_writer can be converted to a `xml2::xml_document` object. An
`xml_fragment` is identical to the output of `xml2::as_list`, so it can
be converted to a `xml2::xml_document` object.

`xml_fragment` supports the following `xml2` methods:

-   `xml2::as_xml_document`, calls the `xml2::read_xml` method
-   `xml2::as_list`, removes the `xml_fragment` class
-   `xml2::write_xml`, writes the xml to a file or console

One can also use the `as_xml_nodeset` function to convert the
`xml_fragment` to a `xml2::xml_nodeset` object.

```{r}
fragment |> xml2::as_xml_document()
fragment |> as_xml_nodeset()
```

`xml_fragment` implements the `xml2::write_xml` method

```{r write_xml, eval=FALSE}
xml2::write_xml(fragment, file="") # print to the console
```

results in:

```{r, echo=FALSE, results='asis'}
cat("```XML\n")
xml2::write_xml(fragment, "")
cat("\n```\n")
```

### Using an `xmlbuilder` object

`xmlbuilder` is an object that allows you to create xml documents in a
feed-forward manner.

With `xml_fragment` one creates the xml document structure as one
(large) R list and then converts it to a xml string or
`xml2::xml_document`, with `xmlbuilder` one incremently builds the xml
document, without having the whole list structure in memory. It also
provides more functions for the output xml document, like adding a
prolog or comment.

It provides the following methods:

-   `start` to start a new element
-   `end` to end the current element
-   `element` to add an element with a value
-   `comment` to add a comment
-   `prolog` to add a prolog
-   `fragment` to add a fragment

```{r xmlbuilder}
library(xmlwriter)

b <- xmlbuilder(allow_fragments = FALSE)
b$comment("This is an xml comment")
b$start("homeless")
  b$start("person", id = "1")
    b$element("name", "John Doe")
    b$element("age", 30)
    b$start("address")
      b$element("street", "123 Main St")
      b$element("city", "Anytown")
      b$element("state", "CA")
      b$element("zip", 12345)
    b$end("address")
  b$end("person")
  b$start("person", id = "2")
    b$element("name", "Jane Doe")
    b$element("age", 25)
    b$start("address")
      b$element("street", "321 Main St")
      b$element("city", "Anytown")
      b$element("state", "CA")
      b$element("zip", 54321)
    b$end("address")
  b$end("person")
  b$fragment(
    person = frag(
      .attr = c(id = "3"),
      name = "Jim Doe",
      age = 35
    )
  )
b$end("homeless")

# includes a xml prolog and comment
b

as.character(b)

# only contains the actual nodes
xml2::as_xml_document(b)
```

# Performance {#performance}

`xmlwriter` is optimized for generating xml documents and fragments from
a R `list` structure that is identical to the `xml2::as_list` output.

```{r performance, eval=TRUE}
library(microbenchmark)

library(xml2)
library(xmlwriter)

# read in a sample 600k xml file as a R list str
doc <- xml2::read_xml("./example/DataGeneric.xml")
doc_list <- xml2::as_list(doc)

# just copy the list and set an extra attribute class="xml_fragment"
# making it a xml_fragment object
doc_fragment <- structure(doc_list, class = "xml_fragment")

# see how long it takes to create an xml document with xml2 and xmlwriter

microbenchmark(
  xml2      = xml2::as_xml_document(doc_list),
  xmlwriter = xml2::as_xml_document(doc_fragment),
  times     = 10
)

```
